---
author: "Michael Koohafkan"
title: "Getting Started with expyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started} 
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document provides a brief introduction to the R package **expyr** 
(**ex**ecute **Py**thon commands from **R**). The package provides the R6 class
`PythonEnv` which provides a bridge between R and a Python server. It is primarily
designed to work with local Python installations, but remote connections are possible.

## Prerequisites

**expyr** works with Python 2.x and 3.x. If you are using Python 3.4 or higher,
everything should work out of the box. If you are using Python 2 or an earlier
version of Python 3, you will need to install the `singledispatch` module as well. 
If you are using Python version 2.7.9 or higher, you can install `singledispatch` 
with `pip` from the command line with 

```
python -m pip install singledispatch
```

If you are using an earlier version of Python, you'll have to 
[install pip](http://pip.readthedocs.io/en/latest/installing/) first.
The easiest way to do this on a Windows machine is to put `get-pip.py` in 
the same directory as `python.exe` and run this command from that  directory:

```
python get-pip.py
```

Then you can install `singledispatch` using `pip`.

## The Python Server

A Python server is provided as a single script (`pysockr.py`) and a couple of 
extensions (discussed in the companion vignette "Extending expyr").

```{r eval = FALSE}
# View the Python server script
file.show("py-src/pysockr.py", package = "expyr")
```

## The PythonEnv Class

The *PythonEnv* class provides an interface to the Python server via socket 
connections and a JSON protocol for transferring variables to/from Python. Each 
instance of the *PythonEnv* class is tied to a unique Python server and 
communicate with independent Python processes. Use the `new()` method to create 
a *PythonEnv* object.

```{r eval = FALSE}
pypath = Sys.which('python')
host = 'localhost'
port = 6011

py = PythonEnv$new(pypath, port, host)
```

Each *PythonEnv* object has a number of bindings that provide information on the
Python environment settings. Most of these bindings can only be changed when the
Python server is stopped.

```{r eval = FALSE}
py$running    # Check if the Python server is running
py$pid        # the process id of the Python server
py$host       # get/set the hostname of the Python server
py$port       # get/set the port number used for communication with the Python server
py$timeout    # get/set the timeout period for listening to the Python server
```

## Interacting with Python 

Before executing Python commands, you must start the Python server using the 
`start()` method. 

```{r eval = FALSE}
py$start()
```

To execute Python code, use the `exec` method:

```{r eval = FALSE}
py$exec("5 + 5")
```

`exec` won't return a value without an explicit `print` statement on the Python
side, and will invisibly retrieve results in R. It can be a little clumsy to
print Python results to the R console:

```{r eval = FALSE}
print(py$exec("print(5 + 5)"))
```

A better option is to use `get`:

```{r eval = FALSE}
py$get("5 + 5")
```

You can assign variables in Python either by executing an assignment statement
with `exec` or by using the `set` command:

```{r eval = FALSE}
py$exec("a = 5")
py$set(a = 5)        # equivalent
```

You can use `get` to retrieve variables too:

```{r eval = FALSE}
py$get('a')
```

To shut down the Python server and close the Python process, 
use the `stop()` method. 

```{r eval = FALSE}
py$stop()
```
